%%--------------------------------------------------------------------------
%%  SECTION 3: MATHEMATICAL THEORY OF DEEP NEURAL NETWORK TRAINING
%%--------------------------------------------------------------------------
\section{Mathematical Theory of Deep Neural Network Training}

\noindent\textit{Why it matters.}\enspace
Deploying neural networks in safety-critical scientific computing demands
more than empirical validation: it requires a rigorous mathematical
understanding of training dynamics and generalization.  Without such theory,
one cannot predict when a neural network will fail---an unacceptable
situation in nuclear safety transport, clinical imaging, or autonomous
systems.  Beyond safety, a quantitative theory of learning dynamics is an
essential design tool: one must understand how a network's inductive bias
interacts with the spectral properties of the target PDE solution to
guarantee convergence and inform architecture choices.

\subsection{The Frequency Principle}

A central theme of my theoretical work is the \emph{Frequency Principle}
(F-principle)~\cite{FP1}: the finding, first systematically documented by
our group, that gradient descent on neural network loss surfaces fits
low-frequency components of the target function first and progressively
captures higher frequencies.  This is \emph{a priori} surprising from an
approximation-theoretic standpoint, and its mathematical explanation reveals
a deep connection between the spectral properties of the neural tangent
kernel and the implicit regularization induced by early stopping.

In~\cite{FP1} we provided the first systematic Fourier analysis
establishing F-principle for two-layer networks, with extensive controlled
experiments spanning regression, classification, and PDE residual
minimization.  Subsequent work established:
\begin{itemize}
  \item A \emph{general theory}~\cite{Luo2021} for deep networks of
    arbitrary width and depth, analyzing the spectral dynamics of the
    infinite-width neural tangent kernel and yielding closed-form predictions
    for training convergence rates.

  \item An \emph{exactly solvable linear ODE model}~\cite{Luo2022} that
    reproduces F-principle dynamics for two-layer networks, enabling analytic
    predictions of convergence as a function of architecture and learning rate
    without any approximation.

  \item \emph{Phase diagrams}~\cite{PD21} characterizing the distinct
    generalization phases of two-layer ReLU networks at the infinite-width
    limit, providing a rigorous map from initialization and architecture to
    implicit bias and test error.
\end{itemize}

These theoretical results carry concrete engineering consequences for the
design of scientific machine learning methods.  The F-principle predicts
that na\"{i}ve PINN formulations will fail to resolve high-frequency
components of multiscale PDE solutions in stiff regimes---exactly the failure
mode that motivates the APNN architecture.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        width=0.85\textwidth,
        height=5cm,
        xmin=0, xmax=2000,
        ymin=0, ymax=1.2,
        xlabel={Training Epochs},
        ylabel={Normalized Error},
        legend style={at={(0.95,0.95)}, anchor=north east, nodes={scale=0.8, transform shape}},
        grid=major,
        cycle list={
          {linkblue, thick},
          {orange, thick, dashed},
          {green!60!black, thick, dashdotted}
        }
      ]
      % Simulation of F-principle training dynamics
      % Low freq drops fast, high freq drops slow
      \addplot[domain=0:2000, samples=50] {exp(-x/200)};
      \addlegendentry{Low Frequency ($k=1$)}

      \addplot[domain=0:2000, samples=50] {0.8*exp(-x/800) + 0.2};
      \addlegendentry{Mid Frequency ($k=4$)}

      \addplot[domain=0:2000, samples=50] {0.9*exp(-x/3000) + 0.1};
      \addlegendentry{High Frequency ($k=8$)}
    \end{axis}
  \end{tikzpicture}
  \vspace{-1.5em}
  \caption{Illustration of the Frequency Principle (F-principle): gradient-based training of neural networks fits target functions sequentially according to frequency, prioritizing low frequencies and exhibiting spectral bias against high frequencies.}
  \label{fig:f-principle}
\end{figure}

Conversely, the exact ODE model
model of~\cite{Luo2022} provides quantitative guidance on how network depth
and width should scale with the regularity of the target PDE solution, turning
previously heuristic design choices into mathematically principled decisions.
