%%--------------------------------------------------------------------------
%%  SECTION 2: KINETIC EQUATIONS & DEEPRTE
%%--------------------------------------------------------------------------
\section{Foundation Models \& Structure-Preserving AI for Kinetic Equations}

\noindent\textit{Why it matters.}\enspace
Kinetic equations are among the most computationally expensive models in all
of science.  A single RTE simulation in nuclear reactor transport involves
$10^8$--$10^{12}$ phase-space unknowns, with run times on classical
supercomputers measured in hours.  Fusion plasma simulations face a starker
challenge: real-time feedback control of a tokamak plasma requires millions
of Vlasov--Maxwell evaluations per second, a throughput that is five to six
orders of magnitude beyond what current particle-in-cell (PIC) codes can
deliver.  Equally important is the \emph{multiscale} structure of these
equations: in the limit $\varepsilon \to 0$ (small Knudsen number), the
kinetic model degenerates to a macroscopic diffusion or fluid system, but
classical numerical discretizations become stiff and impose a time-step
restriction $\Delta t = O(\varepsilon)$, adding an entire order of
computational difficulty to an already intractable problem.

\subsection{DeepRTE: A Foundation Model for Radiative Transfer}

\emph{My most recent and important work} is the development of
\textbf{DeepRTE}~\cite{DeepRTE2026}, the first pre-trained, attention-based
\emph{neural operator foundation model} for the steady-state RTE.  Inspired
by the large language model paradigm, the idea is to train a single model
once on a massive, systematically generated dataset of RTE solutions---covering
the full range of physically relevant scattering anisotropy parameters
$g \in [0.1, 0.8]$---and then deploy it for instant inference on any new
optical geometry without retraining.  DeepRTE learns the solution operator
$\mathcal{A}: (I^{-};\, \mu_t, \mu_s, p) \mapsto I$, mapping the entire
spatial configuration of inflow boundaries, attenuation/scattering
coefficients, and phase functions directly to the full radiation field.

\begin{figure}[H]
  \centering
  \small\sffamily
  \rowcolors{2}{sectionbg}{white}
  \begin{tabularx}{0.98\textwidth}{
      >{\hsize=1.2\hsize\raggedright\arraybackslash}X
      >{\hsize=0.9\hsize\raggedright\arraybackslash}X
      >{\hsize=0.9\hsize\raggedright\arraybackslash}X
    }
    \toprule
    \rowcolor{accentcolor!10}
    \textbf{\textcolor{accentcolor}{Aspect}} & \textbf{\textcolor{accentcolor}{Classical Solvers (e.g., $S_N$)}} & \textbf{\textcolor{accentcolor}{DeepRTE Foundation Model}} \\
    \textbf{Run Time (Inference)} & $10^4$--$10^6$ seconds per geometry & $\sim 10^{-2}$ s (GPU accel., $\sim 10^5\times$ speedup) \\
    \textbf{Mathematical Scope} & Equation discretization & Direct operator mapping $\mathcal{A}: (I^{-}, \mu) \mapsto I$ \\
    \textbf{Reusability} & None natively & Parameterized across optical $g \in [0.1, 0.8]$ \\
    \textbf{Distribution} & Specialized Fortran/C++ libraries & Hugging Face checkpoints, JAX/Flax APIs \\
    \bottomrule
  \end{tabularx}
  \vspace{-0.5em}
  \caption{A paradigm shift in radiative transfer: DeepRTE replaces costly per-instance simulation with instant inference across variable geometries.}
\end{figure}

The engineering realization is as critical as the mathematical novelty.  DeepRTE
is implemented entirely in JAX/Flax for maximum hardware acceleration,
containerized with Docker for reproducible deployment, and distributed via:
\begin{itemize}
  \item GitHub package with full training/inference pipeline:\\\texttt{\href{https://github.com/mazhengcn/deeprte}{github.com/mazhengcn/deeprte}}
  \item Pre-trained checkpoints and datasets on Hugging Face:\\\texttt{\href{https://huggingface.co/mazhengcn/deeprte}{huggingface.co/mazhengcn/deeprte}}
\end{itemize}
This represents a qualitative shift in computational physics: rather than an
isolated scheme for an individual problem, DeepRTE provides the community
with a \emph{reusable computational asset} that amortizes the immense initial
training cost across thousands of downstream, domain-specific tasks.

\subsection{Asymptotic-Preserving Neural Networks}

The central contribution of my group in this area is the development of
\emph{Asymptotic-Preserving Neural Networks}
(APNNs)~\cite{APNN2023,APNN2024}: a fundamentally new class of
physics-informed neural network architectures that provably inherit the
asymptotic-preserving (AP) property of the best classical schemes.  The
essential novelty is architectural.  We encode the micro--macro decomposition
$f = \rho M + \varepsilon g$ (where $M$ is the local Maxwellian equilibrium
and $g$ is the microscopic deviation) directly into the network's structural
inductive bias, rather than including it as a soft loss-function term that
can be violated during optimization.  The ``macro'' branch of the network is
constrained to satisfy the correct limiting diffusion equation as
$\varepsilon \to 0$, and the ``micro'' branch is constrained to vanish in the
same limit.  This is, to our knowledge, the first class of neural network
methods for kinetic equations with a rigorous, verifiable AP guarantee
independent of training quality or random seed.

The framework has been applied across the canonical multiscale kinetic
systems:
\begin{itemize}
  \item \emph{Linear transport equations}~\cite{APNN2023,APNN2024}: uniform
    accuracy from the kinetic regime to the diffusion limit without mesh
    refinement, on problems where standard PINNs fail completely for
    $\varepsilon < 10^{-3}$.

  \item \emph{Vlasov--Poisson--Fokker--Planck system in the high-field
    regime}~\cite{APNN-VPFP2024}: AP behavior in the presence of long-range
    electromagnetic self-interactions, extending the framework to
    nonlinearly coupled systems.

  \item \emph{Gray radiative transfer via even--odd decomposition}
    (AP-CON)~\cite{APCON2024}: operator-learning variants using convolutional
    DeepONets that generalize across $10^4$ distinct optical parameter
    configurations without retraining, achieving accuracy competitive with
    reference $S_N$ solvers while running $10^3\times$ faster at inference.
\end{itemize}

Reference implementations with training scripts, pre-computed datasets, and
documented APIs are publicly available:
\begin{itemize}[nosep, topsep=0.25em, leftmargin=2.5em]
  \item \texttt{\href{https://github.com/mazhengcn/apnns}{github.com/mazhengcn/apnns}}\enspace(APNNs)
  \item \texttt{\href{https://github.com/mazhengcn/apcons}{github.com/mazhengcn/apcons}}\enspace(AP-CON)
\end{itemize}

\subsection{Asymptotic-Preserving Random Feature Methods}

Beyond neural networks, in collaboration with Jingrun~Chen (USTC), we
developed an \emph{asymptotic-preserving random feature method}
(AP-RFM)~\cite{AP-RFM2025} for multiscale radiative transfer.  Random
feature methods expand the solution in a random basis
$\sigma(\omega_j \cdot x + b_j)$ with fixed random weights, offering
mesh-free approximation with theoretically controllable error at
$O(N)$ complexity.  By encoding the micro--macro decomposition into both
the basis construction and the least-squares fitting, AP-RFM achieves uniform
accuracy across all Knudsen numbers at a cost substantially lower than neural
network training.  This result demonstrates that the AP design principle is
a general mathematical framework---not a property specific to any particular
machine learning architecture.
